---
title: "Power Calculation"
author: "Abby Hahs, Ansel Alldredge, Otto Schmidt"
output: pdf_document
date: "2025-05-03"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(simr)
library(ggplot2)
library(gridExtra)
```


#Simulation

```{r}
# data
results <- read_csv("/home/rstudio/users/hahs1/CIR_2024_25_Fish_Vertebrae/results/youngs-modulus/2025-04-30/choices.csv")
length_data <- read.csv("/home/rstudio/users/hahs1/CIR_2024_25_Fish_Vertebrae/data/Perch study - MAIE Lab - Tally.csv")

# filter results to contain only the fish that are not marked inconclusive and are the species PF, split the name to into different sections to differentiate the type (PF), number (01-21), bone number (1-4), and fish name (PF #) 
results_good <- results |>
  dplyr::filter(!inconclusive) %>%
  mutate(
    fish_type = str_sub(name, 1, 2),
    fish_num = str_sub(name, 3, 4),
    bone_type = str_sub(name, 5, 6),
    bone_num = str_sub(name, 7, 8),
    fish_name = str_sub(name, 1, 4)
  ) |>
  dplyr::filter(fish_type != "SV")

# change the fish to be uppercase to match the name format in results_good
length_clean <- length_data |>
  mutate(Individual = toupper(Individual))

# merge results good with length clean. Keep all rows from results_good even if a length is not found. There should be a length value for every fish
results_final <- merge(results_good, length_clean, by.x = "fish_name", by.y = "Individual", all.x = T) 
```


Generate simulated HEIGHT for n.trees per each of n.transects. (Total sample size is n.trees*n.transects)
mu is the overall mean value (assumed to be the same for all transects)
sd.between is between transects
sd.within is within transects

tree = bones
transects = fish

```{r}
mu <- mean(results_final$slope)

# Step 1: compute the mean HEIGHT for each transect
fish_means <- tapply(results_final$slope, results_final$fish_name, mean)

# Step 2: compute the standard deviation of those means
sd.between <- sd(fish_means)

# Step 1: compute the standard deviation within each transect
within_sds <- tapply(results_final$slope, results_final$fish_name, sd)

# Step 2: take the mean of these SDs (or use pooled SD, see note below)
sd.within <- mean(within_sds)


simulate.bones <- function(n.bones,
                           n.fish,
                           bone.type.mus,
                           bone.type.sds,
                           mu = 133.77,
                           sd.between = 30.03,
                           # sd.within = 57.84,
                           bone.types = c("UT", "MT", "LT", "CP")) {
  
  # assign bone types with same proportions as original data
  bone_type_props <- prop.table(table(results_good$bone_type))
  bone.type.assignments <- sample(names(bone_type_props), n.bones, replace = TRUE, prob = bone_type_props)  
  
  bones.sim <- data.frame(matrix(NA, n.bones, n.fish))
  colnames(bones.sim) <- paste("fish", 1:n.fish, sep = "")
  
  fish.mu <- rnorm(n.fish, mu, sd.between)
  
  for(i in 1:n.fish) {
    means <- fish.mu[i] + bone.type.mus[bone.type.assignments]
    # instead of within-fish sd, we use bone-type sd.
    sds <- bone.type.sds[bone.type.assignments]
    
    bones.sim[, i] <- rnorm(n.bones, means, sds)    
  }
  
  # colnames(bones.sim) <- paste("fish", 1:n.fish, sep = "")
 
   bones.sim %>%
    mutate(bone_type = bone.type.assignments,
           id = 1:n.bones) %>%
    pivot_longer(cols = starts_with("fish"),
                 names_to = "fish_name",
                 values_to = "E") %>%
    arrange(fish_name)
}


# simulate.bones <- simulate.bones(235, 21)

# bone-specific E deviation from mean. is added to overall fish mean
bone_type_mu_deviations <- sapply(split(results_good$slope, results_good$bone_type), mean) - mu
# sds for each bone replaces within_sd
bone_type_sds <- tapply(results_final$slope, results_final$bone_type, sd)

simulate.bones <- simulate.bones(235, 21, bone_type_mu_deviations, bone_type_sds)
```

Evaluating results of simulation. Looks good? Yes
```{r}
summary(simulate.bones$E)
summary(results_good$slope)

sd(simulate.bones$E)
sd(results_good$slope)

grid.arrange(
 ggplot(results_good, aes(x = slope)) + 
  geom_boxplot() +
   xlim(c(0, 400)) +
  facet_wrap(~ bone_type),

ggplot(simulate.bones, aes(x = E)) + 
  geom_boxplot() + 
  xlim(c(0, 400)) +
  facet_wrap(~ bone_type) ,
nrow = 2
)

ggplot(bind_rows(
  results_good %>% dplyt::select(slope) %>% mutate(from = "original"),
  simulate.bones %>% dplyr::select(E) %>% mutate(from = "simulated", slope = E)
), aes(x = slope, , fill = from)) +
  geom_density(alpha = 0.5) +
  theme_minimal()
```

All of this seems unnecessary since we are using the simr package below.



# Power Calculation

```{r}
m1 <- lmer(slope ~  bone_type + (1 | fish_name), data = results_good, REML = TRUE)

power_analysis <- powerSim(m1, seed = 123)

summary(power_analysis) # 65% power at our current sample size (21)
```

# increasing sample size 

```{r}
m1.ext <- extend(m1, within = "fish_name", n = 50) # 50 fish. Assumes all other variation stays the same. Seems right, unless we can be more selective with more fish, i.e. have the same number of bones per type.
power_analysis_ext <- powerSim(m1.ext)
summary(power_analysis_ext) # 99% power.
```

## a range of sample sizes

```{r}
plot(powerCurve(m1.ext, within="fish_name", breaks=c(20, 25, 30, 35, 40)))
```




