---
title: "Statistical Analysis Outline"
author: "Abby Hahs, Ansel Alldredge, Otto Schmidt"
output: pdf_document
date: "2025-02-13"
editor_options: 
  chunk_output_type: console
---

This rmd provides a walk through of how the statistical analysis was completed for the Youngs Modulus data.

# Set up and Data Cleaning

Load in the required packages and set up document to obtain data from the directory.

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(gridExtra)
library(ggtext)
library(segmented)
library(changepoint)
library(matrixStats)
library(scales)
library(stats)
library(readr)
library(lme4)
library(HLMdiag)
library(ggResidpanel)
library(mosaic)
library(kableExtra)

#setwd("/home/rstudio/users/hahs1/CIR_2024_25_Fish_Vertebrae")
#data_directory <- "./data"
#script_directory <- "./src/script/"
```

results = the latest results data
length_data = fish length data from Takashi

Need to make sure file path is accurate.

```{r}
# data
results <- read_csv("/home/rstudio/users/hahs1/CIR_2024_25_Fish_Vertebrae/results/youngs-modulus/2025-04-30/choices.csv")
length_data <- read.csv("/home/rstudio/users/hahs1/CIR_2024_25_Fish_Vertebrae/data/Perch study - MAIE Lab - Tally.csv")

# filter results to contain only the fish that are not marked inconclusive and are the species PF, split the name to into different sections to differentiate the type (PF), number (01-21), bone number (1-4), and fish name (PF #) 
results_good <- results |>
  dplyr::filter(!inconclusive) %>%
  mutate(
    fish_type = str_sub(name, 1, 2),
    fish_num = str_sub(name, 3, 4),
    bone_type = str_sub(name, 5, 6),
    bone_num = str_sub(name, 7, 8),
    fish_name = str_sub(name, 1, 4)
  ) |>
  dplyr::filter(fish_type != "SV")

# change the fish to be uppercase to match the name format in results_good
length_clean <- length_data |>
  mutate(Individual = toupper(Individual))

# merge results good with length clean. Keep all rows from results_good even if a length is not found. There should be a length value for every fish
results_final <- merge(results_good, length_clean, by.x = "fish_name", by.y = "Individual", all.x = T) 
```

## Summary Stats

Summary statistics allow us to gain an initial peek into the data and identify possible trends.

```{r}
favstats(slope ~ bone_type, data = results_final) |>
  kable(caption = "Youngs Modulus by Bone Type")
```

## Visualizations

* Studying the E value for each bone. There will be variability both within and between fish which will be addressed in a multilevel linear model. The most useful visualizations to capture these relationships are scatter plots. 

*Within Fish Variability*

Since most fish have > 1 observation for each bone type, checking the within fish variability determines if the observations for each fish are similar or different. Checking this helps determine if any of the values look "off" which would result in checking the algorithm or if the use of a multilevel model is justified.

```{r}
results_final |>
  # order x axis by fish length
  mutate(fish_name = fct_reorder(as.factor(fish_name), Length..cm.)) |>
  ggplot(aes(x = Length..cm., y = slope, color = fish_name, group = fish_name)) +
  # add lines for various y intercepts
  geom_hline(yintercept = 125, color = "gray", size = 0.25) + 
  geom_hline(yintercept = 250, color = "gray", size = 0.25) + 
  geom_hline(yintercept = 375, color = "gray", size = 0.25) +
  # scatterplot with a line connecting the points for each fish
  geom_point(size = 3, alpha = 0.85, show.legend = FALSE) +
  geom_line(size = 1, alpha = 0.75, show.legend = FALSE) +
  labs(x = "Fish Length (cm)", y = "Young's Modulus (N/m^2)", color = "Fish\nName") +
  # theme and formatting
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1
  )) +
  scale_y_continuous(breaks = c(0, 250, 500), limits = c(0, 500)) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 22),
    axis.title.y = element_text(size = 20),
    axis.text.y = element_text(size = 22),
    axis.text.x = element_text(size = 20),
    strip.text = element_text(face = "bold", size = 20),
    panel.grid = element_blank(),
    panel.border = element_rect(fill = NA, color = "gray")
  ) +
  guides(color = guide_legend(nrow = 2)) +
  facet_wrap( ~ fct_relevel(bone_type, "UT", "MT", "LT", "CP"))
```

This plot depicts the variability within fish for E values. The x axis is the fish length, the y axis is the E value, and each fish has a different color.
Generally, the UT and CP bones have more variation than the MT and LT bones. There is also evidence of E values increasing with length for MT, LT, and CP bones. E values for UT do not seem to have a relationship with length.

*Between Fish Variability*

The between fish variability is useful in that we can easily compare between fish and catch discrepancies in E values or expected shape.

```{r}
results_final %>% 
  mutate(xlabel = ifelse(as.numeric(fish_num) >= 8 & fish_num <= 14, bone_type, NA)) %>% 
  # order each plot by fish length
  mutate(fish_name = fct_reorder(as.factor(fish_name), Length..cm.)) |>
ggplot(aes(x = fct_relevel(bone_type, "UT", "MT", "LT", "CP"), y = slope, color = fish_name)) +
  # add lines for chosen y intercepts
  geom_hline(yintercept = 125, color = "gray", size = 0.25) + 
  geom_hline(yintercept = 250, color = "gray", size = 0.25) + 
  geom_hline(yintercept = 375, color = "gray", size = 0.25) + 
  # scatterplot with a smooth line for the average
  geom_point(size = 3, show.legend = FALSE) +
  geom_smooth(
    aes(group = fish_name),
    se = F,
    size = 0.75,
    show.legend = FALSE
  ) +
  # create one plot for each fish
  facet_wrap( ~ fish_name,
              nrow = 3,
              labeller = labeller(fish_name = toupper)) +
  labs(x = "Bone Type", y = "Young's Modulus (N/m^2)") +
 scale_y_continuous(
    breaks = c(0, 250, 500)
  ) +
  # theme and formatting
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 22),
    axis.title.y = element_text(size = 22),
    axis.text.y = element_text(size = 20),
    axis.text.x = element_text(size = 20),
    panel.spacing = unit(0.2, "lines"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_rect(fill = NA, color = "gray"),
    strip.text = element_text(size = 22, face = "bold")
  )
```

This plot depicts the variability between fish for E vales. The x axis is bone type, ordered by location in the spine (UT, MT, LT, CP), the y axis is E, the fish are ordered according to length and the colors correspond to the same fish in within fish variability. We expect to see a "U" shape with higher UT and CP values compared to MT and LT. About 50% of the fish display this trend, while others show no trend or the opposite trend, thus indicating variability between fish.

## Statistical Analysis

* Statistical Analysis focuses on determining if the bone type (UT, MT, LT, CP) is a significant predictor of E value.

*Initial Statistical Analysis*

* An Analysis of Variance test determines if there is an initial association between bone type and E value before adding effects of the individual fish.
* Tukey test looks at each bone type and determines if there are significant differences comparatively. 

```{r}
a1 <- aov(slope~bone_type, data = results_final)
summary(a1) 
```

* We do not have significant evidence that bone type is associated with Youngs Modulus (p=0.105).

```{r}
TukeyHSD(a1)
```

* There is almost a significant difference in Youngs Modulus values for UT compared to MT bones (p=0.06).

*Multilevel model*

A multilevel model allows us to introduce a random intercept for each fish and take into account that each fish is different.

* We first create a model with only the intercept (m0) and then compare it to the model where we add bone type as a predictor (m1). An ANOVA test tells us if the model with bone type is significantly better at predicting E values than the model with only the intercept.

```{r}
m0 <- lmer(slope ~  1 + (1 | fish_name), data = results_final, REML = TRUE)

m1 <- lmer(slope ~  bone_type + (1 | fish_name), data = results_final, REML = TRUE)

anova(m1, m0)
```

Interpretation: The model with the bone type is significantly better at predicting the E values than the model with just the intercept (p=0.04).

*Residuals*

We check residuals to make sure the model fit is good and to see if any of the variables need to be transformed.

```{r}
resid_panel(m1)
```

Interpretation: The histogram of residuals is right skewed and the QQ plot is not linear at the right end. As a result, we will log transform the response variable(E). 

*log(E)*

```{r}
m00 <- lmer(I(log(slope)) ~  1 + (1 | fish_name), data = results_final, REML = TRUE)

m11 <- lmer(I(log(slope)) ~  bone_type + (1 | fish_name), data = results_final, REML = TRUE)
summary(m11)

anova(m11, m00)
```

Interpretation: The log transformed model with the bone type is significantly better at predicting the E values than the log transformed model with just the intercept (p=0.005).

```{r}
resid_panel(m11)
```

Interpretation: The histogram of the residuals is a normal distribution and the QQ plot is linear. The log-transformation successfully enhanced the model.

*Changing the reference level *

By changing the reference level from the default (CP), we can see if there are significant differences when compared to a different bone type.

```{r}
# create new data and manually relevel the bone types
decisions_relevel_UT <- results_final |>
   mutate(bone_type = fct_relevel(bone_type, "UT", "CP", "LT", "MT"))

# run the model
m2.ut <- lmer(I(log(slope)) ~  relevel(bone_type, ref = "UT") + (1 | fish_name), data = decisions_relevel_UT, REML = TRUE)

summary(m2.ut)
```

Interpretation: The E values for the UT bones are significantly greater than the values for the other bone types as seen by t values > |2|. Since E is logged, we can say that the MT bones have values 24.3% lower than UT ($1-e^{-0.27914}$), LT bones are 18.8% lower ($1-e^{-0.20917}$), and CP bones are 16.3% lower than UT ($1 - e^{-0.17852}$). 

We can continue changing the reference level.

```{r}
decisions_relevel_LT <- results_final |>
   mutate(bone_type = fct_relevel(bone_type, "LT", "CP", "MT", "UT"))

m2.lt <- lmer(slope ~  relevel(bone_type, ref = "LT") + (1 | fish_name), data = decisions_relevel_LT, REML = TRUE)

summary(m2.lt)

decisions_relevel_MT <- results_final |>
   mutate(bone_type = fct_relevel(bone_type, "MT", "CP", "LT", "UT"))

m2.mt <- lmer(slope ~  relevel(bone_type, ref = "MT") + (1 | fish_name), data = decisions_relevel_MT, REML = TRUE)

summary(m2.mt)
```

Interpretation: All t values for bones not including UT were <= |2|, thus indicating that there are no significant differences between any of the other bone pairs.