---
title: "Statistical Analysis Outline"
author: "Abby Hahs, Ansel Alldredge, Otto Schmidt"
output: pdf_document
date: "2025-02-13"
editor_options: 
  chunk_output_type: console
---

# Set up and Data Cleaning

Load in the required packages and set up document to obtain data from the directory.

```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(ggtext)
library(segmented)
library(changepoint)
library(matrixStats)
library(scales)
library(stats)
library(readr)
library(lme4)
library(HLMdiag)
library(ggResidpanel)

setwd("/home/rstudio/users/hahs1/CIR_2024_25_Fish_Vertebrae")
data_directory <- "./data"
script_directory <- "./src/script/"
```

results = the latest results data
length_data = fish length data from Takashi

```{r}
results <- read_csv("results/youngs-modulus/2025-04-17/choices.csv")
length_data <- read.csv("data/Perch study - MAIE Lab - Tally.csv")

results_good <- results |>
  dplyr::filter(!inconclusive) %>%
  mutate(
    fish_type = str_sub(name, 1, 2),
    fish_num = str_sub(name, 3, 4),
    bone_type = str_sub(name, 5, 6),
    bone_num = str_sub(name, 7, 8),
    fish_name = str_sub(name, 1, 4)
  ) |>
  dplyr::filter(fish_type != "SV")

length_clean <- length_data |>
  mutate(Individual = toupper(Individual))

results_final <- merge(results_good, length_clean, by.x = "fish_name", by.y = "Individual", all.x = T) 
```

## Visualizations

* Studying the E value for each bone. There will be variability both within and between fish which will be addressed in a multilevel linear model. The most useful visualizations to capture these relationships are scatterplots. 

*Youngs Modulus vs fish length*

```{r}
p1 <- results_final |>
  # filter to contain only one bone type
  filter(bone_type == "UT") |>
  # plot the fish length and slope
  ggplot(aes(x = Length..cm., y = slope)) + 
  # give each fish a color
  geom_point(aes(color = fish_name)) +
  # approximate with a linear slope
  geom_smooth(method = lm, se = F) +
  # set the y axis limits so each plot matches
  scale_y_continuous(limits = c(0,500)) +
  labs(title = "UT")

p2 <- results_final |>
  filter(bone_type == "MT") |>
  ggplot(aes(x = Length..cm., y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "MT") +
  scale_y_continuous(limits = c(0,500))

p3 <- results_final |>
  filter(bone_type == "LT") |>
  ggplot(aes(x = Length..cm., y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "LT") +
  scale_y_continuous(limits = c(0,500))

p4 <- results_final |>
  filter(bone_type == "CP") |>
  ggplot(aes(x = Length..cm., y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "CP") +
  scale_y_continuous(limits = c(0,550))

grid.arrange(p1,p2,p3,p4)
```

Here we see that for MT, LT, and CP bones as the fish length increases, so does the E value.

*Within Fish Variability*

```{r}
results_final |>
  # order x axis by fish length
  mutate(fish_name = fct_reorder(as.factor(fish_name), Length..cm.)) |>
  ggplot(aes(x = Length..cm., y = slope, color = fish_name, group = fish_name)) +
  # add lines for various y intercepts
  geom_hline(yintercept = 125, color = "gray", size = 0.25) + 
  geom_hline(yintercept = 250, color = "gray", size = 0.25) + 
  geom_hline(yintercept = 375, color = "gray", size = 0.25) +
  # scatterplot with a line connecting the points for each fish
  geom_point(size = 3, alpha = 0.85, show.legend = FALSE) +
  geom_line(size = 1, alpha = 0.75, show.legend = FALSE) +
  labs(x = "Fish Length (cm)", y = "Young's Modulus (N/m^2)", color = "Fish\nName") +
  # theme and formatting
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1
  )) +
  scale_y_continuous(breaks = c(0, 250, 500), limits = c(0, 500)) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 22),
    axis.title.y = element_text(size = 20),
    axis.text.y = element_text(size = 22),
    axis.text.x = element_text(size = 20),
    strip.text = element_text(face = "bold", size = 20),
    panel.grid = element_blank(),
    panel.border = element_rect(fill = NA, color = "gray")
  ) +
  guides(color = guide_legend(nrow = 2)) +
  facet_wrap( ~ fct_relevel(bone_type, "UT", "MT", "LT", "CP"))
```

This plot depicts the variability within fish for E values. The UT and CP bones appear to have more variation than the MT and LT bones. 

*Between Fish Variability*

```{r}
results_final %>% 
  mutate(xlabel = ifelse(as.numeric(fish_num) >= 8 & fish_num <= 14, bone_type, NA)) %>% 
  # order each plot by fish length
  mutate(fish_name = fct_reorder(as.factor(fish_name), Length..cm.)) |>
ggplot(aes(x = fct_relevel(bone_type, "UT", "MT", "LT", "CP"), y = slope, color = fish_name)) +
  # add lines for chosen y intercepts
  geom_hline(yintercept = 125, color = "gray", size = 0.25) + 
  geom_hline(yintercept = 250, color = "gray", size = 0.25) + 
  geom_hline(yintercept = 375, color = "gray", size = 0.25) + 
  # scatterplot with a smooth line for the average
  geom_point(size = 3, show.legend = FALSE) +
  geom_smooth(
    aes(group = fish_name),
    se = F,
    size = 0.75,
    show.legend = FALSE
  ) +
  # create one plot for each fish
  facet_wrap( ~ fish_name,
              nrow = 3,
              labeller = labeller(fish_name = toupper)) +
  labs(x = "Bone Type", y = "Young's Modulus (N/m^2)") +
 scale_y_continuous(
    breaks = c(0, 250, 500)
  ) +
  # theme and formatting
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 22),
    axis.title.y = element_text(size = 22),
    axis.text.y = element_text(size = 20),
    axis.text.x = element_text(size = 20),
    panel.spacing = unit(0.2, "lines"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_rect(fill = NA, color = "gray"),
    strip.text = element_text(size = 22, face = "bold")
  )
```

In this plot we can see the E values for each fish and compare the curves. A "U" shape is expected. We can see that the shapes between fish are very different indicating variation.

## Statistical Analysis

* Statistical Analysis focuses on determining if the bone type (UT, MT, LT, CP) is a significant predictor of E value.

*Initial Statistical Analysis*

* Analysis of Variance test determines if there is an initial association between bone type and E value before adding effects of the individual fish.
* Tukey test looks at each bone type and determines if there are significant differences comparatively. 

```{r}
a1 <- aov(slope~bone_type, data = results_final)
summary(a1) 
```

* We have significant evidence that bone type is associated with Youngs Modulus (p<0.001).

```{r}
TukeyHSD(a1)
```

* There is a significant difference in Youngs Modulus values for UT compared to LT bones (p=0.03) and UT compared to MT bones (p<0.001).

*Multilevel model*

A multilevel model allows us to introduce a random intercept for each fish and take into account that each fish is different.

* We first create a model with only the intercept (m0) and then compare it to the model where we add bone type as a predictor (m1). An ANOVA test tells us if the model with bone type is significantly better at predicting E values.

```{r}
m0 <- lmer(slope ~  1 + (1 | fish_name), data = results_final, REML = TRUE)

m1 <- lmer(slope ~  bone_type + (1 | fish_name), data = results_final, REML = TRUE)

anova(m1, m0)
```

Interpretation: The model with the bone type ....

*Residuals*

```{r}
resid_panel(m1)
## In general, residuals are pretty good. QQ plot is bad on right end.
## Data is a little right skewed?

# OUTLIERS

# Add columns for ID and Pearson residuals
decisions_resid <- decisions_length %>%
  mutate(Pearson_resids = residuals(m1, type = "pearson"),
         id = row_number())

# create smaller data set with points we want to label
outliers <- decisions_resid %>%
  filter(Pearson_resids > 2 | Pearson_resids < (-2))

ggplot(data = decisions_resid, mapping = aes(x = id, y = Pearson_resids)) +
  geom_point(alpha = 1/4) +
  geom_hline(yintercept = c(-2, 2), color = "red") +
  geom_point(size = 3, shape = 1, data = outliers) +
  geom_label_repel(aes(label = name, colour = name), 
    data = outliers, show.legend = FALSE)
```

*log(YM)*

```{r}
m11 <- lmer(I(log(slope)) ~  bone_type + (1 | fish_name), data = decisions_length, REML = TRUE)
summary(m11)

m00 <- lmer(I(log(slope)) ~  1 + (1 | fish_name), data = decisions_length, REML = TRUE)
anova(m11, m00) # p=0.0007

resid_panel(m11)
# this looks better

# OUTLIERS

# Add columns for ID and Pearson residuals
decisions_resid2 <- decisions_length %>%
  mutate(Pearson_resids = residuals(m11, type = "pearson"),
         id = row_number())

# create smaller data set with points we want to label
outliers2 <- decisions_resid2 %>%
  filter(Pearson_resids > 2 | Pearson_resids < (-2))

ggplot(data = decisions_resid2, mapping = aes(x = id, y = Pearson_resids)) +
  geom_point(alpha = 1/4) +
  geom_hline(yintercept = c(-2, 2), color = "red") +
  geom_point(size = 3, shape = 1, data = outliers2) +
  geom_label_repel(aes(label = name, colour = name), 
    data = outliers2, show.legend = FALSE)
```


*multilevel models with UT as reference level*

```{r}
decisions_relevel_UT <- decisions_good |>
   mutate(bone_type = fct_relevel(bone_type, "UT", "CP", "LT", "MT"))

m2.ut <- lmer(I(log(slope)) ~  relevel(bone_type, ref = "UT") + (1 | fish_name), data = decisions_relevel_UT, REML = TRUE)

summary(m2.ut) # differences between UT-MT and UT-LT are significant
```

# Other Analysis

*multilevel changing reference level*

```{r}
decisions_relevel_CP <- results_good |> 
   mutate(bone_type = fct_relevel(bone_type, "CP", "LT", "MT", "UT"))

m2.cp <- lmer(log(slope) ~  relevel(bone_type, ref = "CP") + (1 | fish_name), data = decisions_relevel_CP, REML = TRUE)

summary(m2.cp)

decisions_relevel_LT <- decisions_good |>
   mutate(bone_type = fct_relevel(bone_type, "LT", "CP", "MT", "UT"))

m2.lt <- lmer(slope ~  relevel(bone_type, ref = "LT") + (1 | fish_name), data = decisions_relevel_LT, REML = TRUE)

summary(m2.lt)

decisions_relevel_MT <- decisions_good |>
   mutate(bone_type = fct_relevel(bone_type, "MT", "CP", "LT", "UT"))

m2.mt <- lmer(slope ~  relevel(bone_type, ref = "MT") + (1 | fish_name), data = decisions_relevel_MT, REML = TRUE)

summary(m2.mt)
```

*group bones by location. (CP and UT) vs (MT and LT)*

```{r}
decisions_by_location <- decisions_good |>
  mutate(location = case_when(bone_type %in% c("CP", "UT") ~ "outer",
        bone_type %in% c("MT", "LT") ~ "inner"))


a5 <- aov(slope ~ location, data = decisions_by_location)
summary(a5)

TukeyHSD(a5)

m0 <- lmer(slope ~  1 + (1 | fish_name), data = decisions_by_location, REML = TRUE)

m4 <- lmer(slope ~  location + (1 | fish_name), data = decisions_by_location, REML = TRUE)
summary(m4) # t = 2.282

anova(m4, m0) # p = 0.02416

# When we group the bones by location, the model with the bone location is significantly better than the model without the bone location (p=0.02416). 

# Is there a problem with doing this since then our data is only inner or outer?
```
