---
title: "statistical_analysis"
author: "Abby Hahs, Ansel Alldredge, Otto Schmidt"
output: pdf_document
date: "2025-02-13"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(coro)
library(gridExtra)
library(ggtext)
library(segmented)
library(changepoint)
library(matrixStats)
library(scales)
library(stats)
library(readr)
library(lme4)
library(HLMdiag)
library(ggResidpanel)
library(ggrepel)

setwd("/home/rstudio/users/hahs1/CIR_2024_25_Fish_Vertebrae")
data_directory <- "./data"
script_directory <- "./src/script/"
```

```{r}
# Load in data and clean it
decisions <- read_csv("results/youngsModulus/decisions.csv")
yieldStress <- read_csv("results/yieldStress/yieldStress.csv")
method <- read_csv("results/youngsModulus/results.csv")
misidentified <- c("pf05UT2", "pf16MT2")
area <- read_csv("data/area.csv")

area_clean <- area |>
  rename(bone_size = `Area (m^2)`,
         segment = `Segment (UT, MT, LT or CP)`,
         trial = `Trial # (at least 01-03)`) |>
  unite("fish_name", Individual, segment, trial, sep ="")

decisions_good <- decisions |>
  mutate(misidentified_flag = if_else(name %in% misidentified, TRUE, FALSE)) |> 
  dplyr::filter(!inconclusive) %>%
  mutate(
    fish_type = str_sub(name, 1, 2),
    fish_num = str_sub(name, 3, 4),
    bone_type = str_sub(name, 5, 6),
    # bone_type = factor relvel todo
    bone_num = str_sub(name, 7, 8),
    fish_name = str_sub(name, 1, 4)
  ) |>
  dplyr::filter(fish_type != "sv")

decisions_area <- merge(decisions_good, area_clean, by.x = "name", by.y = "fish_name")


length_data <- read.csv(("Perch study - MAIE Lab - Tally.csv"))

decisions_length <- merge(decisions_good, length_data, by.x = "fish_name", by.y = "Individual", all.x = T) 
```

## Visualizations - Usefull

```{r}
decisions_good |> # rem this bone order is flipped
  ggplot(aes(x = bone_type, y = slope)) + 
  geom_boxplot() 
```

*YM vs fish length*

```{r}
p1 <- decisions_length |>
  filter(bone_type == "UT") |>
  ggplot(aes(x = Length..cm., y = slope)) + 
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  scale_y_continuous(limits = c(0,500)) +
  labs(title = "UT")

p2 <- decisions_length |>
  filter(bone_type == "MT") |>
  ggplot(aes(x = Length..cm., y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "MT") +
  scale_y_continuous(limits = c(0,500))

p3 <- decisions_length |>
  filter(bone_type == "LT") |>
  ggplot(aes(x = Length..cm., y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "LT") +
  scale_y_continuous(limits = c(0,500))

p4 <- decisions_length |>
  filter(bone_type == "CP") |>
  ggplot(aes(x = Length..cm., y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "CP") +
  scale_y_continuous(limits = c(0,550))

grid.arrange(p1,p2,p3,p4)
```


*Within fish variability*

```{r}
# color the 10 misidentified bones
#decisions_length <- decisions_length |>
 # mutate(color_group = ifelse(misidentified_flag, name, "Good"))

# Order x axis by length

# Plot with unique colors for highlighted fish
p1 <- decisions_length |>
  filter(bone_type == "CP") |>
  mutate(fish_name = fct_reorder(fish_name, Length..cm.)) |>
  ggplot(aes(x = fish_name, y = slope)) +
  geom_point(size = 3, alpha = 0.5) +
  # scale_color_manual(values = c(setNames(rainbow(15), color_group), "Good" = "black")) +
  labs(title = "CP", x = "Fish Name", y = "Young's Modulus") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  scale_y_continuous(limits = c(0, 600))

p2 <- decisions_length |>
  filter(bone_type == "LT") |>
  mutate(fish_name = fct_reorder(fish_name, Length..cm.)) |>
  ggplot(aes(x = fish_name, y = slope)) +
  geom_point(size = 3, alpha = 0.5) +
  # scale_color_manual(values = c(setNames(rainbow(15), misidentified_flag), "Good" = "black")) +
  labs(title = "LT", x = "Fish Name", y = "Young's Modulus") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  scale_y_continuous(limits = c(0, 600))

p3 <- decisions_length |>
  filter(bone_type == "MT") |>
  mutate(fish_name = fct_reorder(fish_name, Length..cm.)) |>
  ggplot(aes(x = fish_name, y = slope)) +
  geom_point(size = 3, alpha = 0.5) +
  # scale_color_manual(values = c(setNames(rainbow(15), misidentified_flag), "Good" = "black")) +
  labs(title = "MT", x = "Fish Name", y = "Young's Modulus") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  scale_y_continuous(limits = c(0, 600))

p4 <- decisions_length |>
  filter(bone_type == "UT") |>
  mutate(fish_name = fct_reorder(fish_name, Length..cm.)) |>
  ggplot(aes(x = fish_name, y = slope)) +
  geom_point(size = 3, alpha = 0.5) +
  # scale_color_manual(values = c(setNames(rainbow(15), ), "Good" = "black")) +
  labs(title = "UT", x = "Fish Name", y = "Young's Modulus") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  scale_y_continuous(limits = c(0, 600))

grid.arrange(p1,p2,p3,p4)
```

*between fish variability*

```{r}
decisions_good |>
  ggplot(aes(x = fct_relevel(bone_type, "UT", "MT", "LT", "CP"), y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(aes(group = fish_name, color = fish_name), se=F) + 
  facet_wrap(~ fish_name) +
  labs(x = "Bone Type", y = "Young's Modulus") +
  theme(legend.position = "none")
```


## Statistical Analysis

*Youngs modulus*

```{r}
a1 <- aov(slope~bone_type, data = decisions_good)
summary(a1)   # p = 0.017

TukeyHSD(a1)
# difference between UT and MT is significant
```

*Multilevel model*

```{r}
# Random intercept for each fish.
m1 <- lmer(slope ~  bone_type + (1 | fish_name), data = decisions_length, REML = TRUE)
summary(m1)

m0 <- lmer(slope ~  1 + (1 | fish_name), data = decisions_length, REML = TRUE)
anova(m1, m0) # p = 0.008
```

*Residuals*

```{r}
resid_panel(m1)
## In general, residuals are pretty good. QQ plot is bad on right end.
## Data is a little right skewed?

# OUTLIERS

# Add columns for ID and Pearson residuals
decisions_resid <- decisions_length %>%
  mutate(Pearson_resids = residuals(m1, type = "pearson"),
         id = row_number())

# create smaller data set with points we want to label
outliers <- decisions_resid %>%
  filter(Pearson_resids > 2 | Pearson_resids < (-2))

ggplot(data = decisions_resid, mapping = aes(x = id, y = Pearson_resids)) +
  geom_point(alpha = 1/4) +
  geom_hline(yintercept = c(-2, 2), color = "red") +
  geom_point(size = 3, shape = 1, data = outliers) +
  geom_label_repel(aes(label = name, colour = name), 
    data = outliers, show.legend = FALSE)
```

*log(YM)*

```{r}
m11 <- lmer(I(log(slope)) ~  bone_type + (1 | fish_name), data = decisions_length, REML = TRUE)
summary(m11)

m00 <- lmer(I(log(slope)) ~  1 + (1 | fish_name), data = decisions_length, REML = TRUE)
anova(m11, m00) # p=0.0007

resid_panel(m11)
# this looks better

# OUTLIERS

# Add columns for ID and Pearson residuals
decisions_resid2 <- decisions_length %>%
  mutate(Pearson_resids = residuals(m11, type = "pearson"),
         id = row_number())

# create smaller data set with points we want to label
outliers2 <- decisions_resid2 %>%
  filter(Pearson_resids > 2 | Pearson_resids < (-2))

ggplot(data = decisions_resid2, mapping = aes(x = id, y = Pearson_resids)) +
  geom_point(alpha = 1/4) +
  geom_hline(yintercept = c(-2, 2), color = "red") +
  geom_point(size = 3, shape = 1, data = outliers2) +
  geom_label_repel(aes(label = name, colour = name), 
    data = outliers2, show.legend = FALSE)
```


*multilevel models with UT as reference level*

```{r}
decisions_relevel_UT <- decisions_good |>
   mutate(bone_type = fct_relevel(bone_type, "UT", "CP", "LT", "MT"))


m2.ut <- lmer(I(log(slope)) ~  relevel(bone_type, ref = "UT") + (1 | fish_name), data = decisions_relevel_UT, REML = TRUE)


summary(m2.ut) # differences between UT-MT and UT-LT are significant
```

## Visualizations - Not as good

*bone number on x axis*

```{r}
decisions_good |>
  filter(bone_type == "CP") |>
  ggplot(aes(x = bone_num, y = slope)) + 
  geom_point(aes(color = as.factor(fish_name)), show.legend = FALSE) + 
  geom_smooth(method ="lm", se = FALSE, aes(group  = fish_name, color = as.factor(fish_name)), size = 0.5, show.legend = FALSE) +
  geom_smooth(method ="lm", se = FALSE) + 
  labs(title = "CP") -> p1

decisions_good |>
  filter(bone_type == "LT") |>
  ggplot(aes(x = bone_num, y = slope)) + 
  geom_point(aes(color = as.factor(fish_name)), show.legend = FALSE) + 
  geom_smooth(method ="lm", se = FALSE, aes(group  = fish_name, color = as.factor(fish_name)), size = 0.5, show.legend = FALSE) +
  geom_smooth(method ="lm", se = FALSE) + 
  labs(title = "LT") -> p2

decisions_good |>
  filter(bone_type == "MT") |>
  ggplot(aes(x = bone_num, y = slope)) + 
  geom_point(aes(color = as.factor(fish_name)), show.legend = FALSE) + 
  geom_smooth(method ="lm", se = FALSE, aes(group  = fish_name, color = as.factor(fish_name)), size = 0.5, show.legend = FALSE) +
  geom_smooth(method ="lm", se = FALSE) + 
  labs(title = "MT") -> p3

decisions_good |>
  filter(bone_type == "UT") |>
  ggplot(aes(x = bone_num, y = slope)) + 
  geom_point(aes(color = as.factor(fish_name)), show.legend = FALSE) + 
  geom_smooth(method ="lm", se = FALSE, aes(group  = fish_name, color = as.factor(fish_name)), size = 0.5, show.legend = FALSE) +
  geom_smooth(method ="lm", se = FALSE) + 
  labs(title = "UT") -> p4

grid.arrange(p1, p2, p3, p4)
```

```{r}
# spaghetti plots of each fish bone
decisions_good |>
  filter(bone_type == "CP") |>
  ggplot(aes(x = bone_num, y = slope)) + 
  geom_line(aes(group = fish_name, color = fish_name)) +
  labs(title = "CP") -> p1

decisions_good |>
  filter(bone_type == "LT") |>
  ggplot(aes(x = bone_num, y = slope)) + 
  geom_line(aes(group = fish_name, color = fish_name)) +
  labs(title = "LT") -> p2

decisions_good |>
  filter(bone_type == "MT") |>
  ggplot(aes(x = bone_num, y = slope)) + 
  geom_line(aes(group = fish_name, color = fish_name)) +
  labs(title = "MT") -> p3

decisions_good |>
  filter(bone_type == "UT") |>
  ggplot(aes(x = bone_num, y = slope)) + 
  geom_line( aes(group = fish_name, color = fish_name)) +
  labs(title = "UT") -> p4

grid.arrange(p1, p2, p3, p4)
```

*YM v bone size*

```{r}
p1 <- decisions_area |>
  filter(bone_type == "CP") |>
  ggplot(aes(x = bone_size, y = slope)) + # bone size
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  scale_y_continuous(limits = c(0,500))
  labs(title = "CP")

p2 <- decisions_area |>
  filter(bone_type == "LT") |>
  ggplot(aes(x = bone_size, y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "LT") +
  scale_y_continuous(limits = c(0,500))

p3 <- decisions_area |>
  filter(bone_type == "MT") |>
  ggplot(aes(x = bone_size, y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "MT") +
  scale_y_continuous(limits = c(0,500))

p4 <- decisions_area |>
  filter(bone_type == "UT") |>
  ggplot(aes(x = bone_size, y = slope)) +
  geom_point(aes(color = fish_name)) +
  geom_smooth(method = lm, se = F) +
  labs(title = "UT") +
  scale_y_continuous(limits = c(0,550))

grid.arrange(p1,p2,p3,p4)
```

## Analysis - not usefull

```{r}
a2 <- aov(slope~fish_name, data = decisions_good) # p < .001. fine; only significant difference is between pf18 and pf07
summary(a2)
TukeyHSD(a2)

# indicates variability between fish
```

*multilevel changing reference level*

```{r}
decisions_relevel_CP <- decisions_good |> 
   mutate(bone_type = fct_relevel(bone_type, "CP", "LT", "MT", "UT"))

decisions_relevel_LT <- decisions_good |>
   mutate(bone_type = fct_relevel(bone_type, "LT", "CP", "MT", "UT"))

decisions_relevel_MT <- decisions_good |>
   mutate(bone_type = fct_relevel(bone_type, "MT", "CP", "LT", "UT"))

m2.cp <- lmer(slope ~  relevel(bone_type, ref = "CP") + (1 | fish_name), data = decisions_relevel_CP, REML = TRUE)

m2.lt <- lmer(slope ~  relevel(bone_type, ref = "LT") + (1 | fish_name), data = decisions_relevel_LT, REML = TRUE)

m2.mt <- lmer(slope ~  relevel(bone_type, ref = "MT") + (1 | fish_name), data = decisions_relevel_MT, REML = TRUE)

summary(m2.cp)
summary(m2.lt)
summary(m2.mt)
```

*group bones by location. (CP and UT) vs (MT and LT)*

```{r}
age_sex_mat <- age_sex[5:8, ]
age_sex_mat <- as.matrix(age_sex_mat)
age_sex_mat <- matrix(age_sex_mat, ncol = 20)

```


```{r}
decisions_by_location <- decisions_good |>
  mutate(location = case_when(bone_type %in% c("CP", "UT") ~ "outer",
        bone_type %in% c("MT", "LT") ~ "inner"))


a5 <- aov(slope ~ location, data = decisions_by_location)
summary(a5)

TukeyHSD(a5)

m0 <- lmer(slope ~  1 + (1 | fish_name), data = decisions_by_location, REML = TRUE)

m4 <- lmer(slope ~  location + (1 | fish_name), data = decisions_by_location, REML = TRUE)
summary(m4) # t = 2.282

anova(m4, m0) # p = 0.02416

# When we group the bones by location, the model with the bone location is significantly better than the model without the bone location (p=0.02416). 

# Is there a problem with doing this since then our data is only inner or outer?
```
